An Effective Model of HL-60 Differentiation

### 2017-04-19 Paper_v2.tex
Paragraph for gene KOs

The signal integration and phenotype modules described ATRA-induced gene expression events in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
which then in turn activated the phenotypic program. In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex,
regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
We estimated the parameters which appeared in the control laws regulating these programs
from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
However, the bulk of the remaining model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
Lastly, we used the \textit{predicted} values of the P21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
The phenotype module predicted P21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and wash robust to uncertainty in the measured data (Fig. \ref{fig:model-cell-cycle-simulations}B, gray).
Taken together, the output of the signal integration and phenotypic modules was consistent with time-series and steady-state measurements, thereby validating the assumed molecular connectivity.
Moreover, outputs from the phenotype module described the trend in ATRA-induced G0 cell cycle arrest.
Next, we explored which nodes and interactions between nodes in the signal integration module most influenced the system response.

@@

>>>>

The signal integration and phenotype modules described ATRA-induced gene expression in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
which then in turn activated the phenotypic program. In particular, Trigger, which is a surrogate for the RAR$\alpha$/RXR transcriptional complex,
regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and EGR1.
In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or P47Phox.
We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
We estimated the parameters which appeared in the control laws regulating these programs
from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
However, the bulk of the remaining model parameters were taken from directly from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
Lastly, we used the \textit{predicted} values of the P21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
The phenotype module predicted P21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and wash robust to uncertainty in the measured data (Fig. \ref{fig:model-cell-cycle-simulations}B, gray).
Taken together, the output of the signal integration and phenotypic modules was consistent with time-series and steady-state measurements, thereby validating the assumed molecular connectivity.
Moreover, outputs from the phenotype module described the trend in ATRA-induced G0 cell cycle arrest.
Next, we explored which nodes and interactions between nodes in the signal integration module most influenced the system response.

@@

The PU.1 and AP1 proteins were critical regulators of the ATRA-induced signal integration and phenotypic programs (Fig. \ref{fig:model-response-ko}).
We conducted pairwise knock simulations of genes in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaF-S621 signals.
The difference between the system state with and without the gene knockouts (encoded in a state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
All simulations were conducted using the best fit parameter set.
The first four modes described $>$99\% of the gene knockout variance,
with the most important components of these modes being the PU.1 and AP1 proteins, and to a much lesser extent GFI1 and CEBPa (Fig. \ref{fig:model-response-ko}A).
To better understand which connections involving the PU.1 and AP1 proteins were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
SVD decomposition of the state displacement matrix assembled from the pairwise deletion of interactions, suggested the first five modes accounted for $>$99\% of the variance resulting from deletion of the interactions.
The most sensitive interactions for the PU.1 protein involved the regulation of P47Phox expression, and to a lesser extent AP1 and EGR1 expression (Fig. \ref{fig:model-response-ko}B).
On the other other, the most sensitive connections for AP1 involved the regulation of P21 expression, and the mutual activation of PU.1 and AP1 expression.
Taken together, these results suggested that the PU.1 and AP1 proteins acted as important regulators for both the signal integration and phenotype modules.
However, this analysis did not given insight into which components of the signal initiation module was important, and particularly the nature of the signalsome complex.
Toward this question, we experimentally explored the composition and regulation of the signalsome complex.
